### 2016-01-31 inputs/materials_and_methods.tex
phrasing

\subsection{Imputation methods}
\label{sec:imputation_methods}
We considered six imputation methods: Mean Value Imputation (MNI), K-Nearest Neighbour Imputation (KNNI), Random Forest Imputation (RFI), Singular Value Decomposition Imputation (SVDI), Localized Haplotype Clustering Imputation (``Beagle'') and the ``FILLIN'' algorithm from the ``Tassel'' programming suite. For all algorithms, we imputed a $m\times n$ matrix of m individuals and n markers whose data points, defined in the set \{0,1,2,NA\}, represented the three possible genotypes (AA, AB, and BB) and the missing value, respectively. 
\paragraph{MNI}
\label{par:MNI}
: in Mean Value Imputation each missing data point was simply replaced by the mean of the non-missing values for that marker, which was then discretized to the closest value in \{0,1,2\}. 
\paragraph{KNNI}
\label{par:KNNI}
: in K-Nearest Neighbors Imputation, missing data points were imputed based on the weighted average of the K closest markers \cite{troyanskaya_missing_2001}. This K-neighbourhood was defined by the Simple Matching Coefficient distance function \cite{schwender_statistical_2007}, specifically designed for categorical data. $K=4$ was used in KNNI, and neighbour values were weighted by the reciprocal of their distance from the data point to be imputed.
\paragraph{SVDI}
\label{par:SVDI}
: Singular Value Decomposition (SVD) imputation was based on the following factorization of the genotype matrix M (m markers, n individuals):

@@

\paragraph{FILLIN}
The ``Fast Inbred Line Library ImputatioN'' (``FILLIN'') is an imputation method optimised for inbred populations implemented in the ``Tassel'' programming suite \cite{swarts_novel_2014}. ``FILLIN'' is based on haplotype reconstruction around recombination break points; haplotypes are then clustered per genotype similarity using the Hamming distance function. This information is eventually used to impute the target locus in an iterative approach that attemps, through a Markov Chain MonteCarlo (MCMC) process, to maximise the likelihood of the observed SNP calls given the unobserved imputed genotypes.

[ADD command line!!]

\subsection{Imputation procedure}
\label{sec:imputation_procedure}
In order to assess the imputation performance of the different methods, we introduced increasing proportions of artificial missing genotypes in the data sets. These were then imputed with the various algorithms, measuring the resulting imputation accuracy and computation time. Unlike imputation experiments with SNP-chips (e.g., \cite{hickey_factors_2012}), GBS data intrinsically present a much larger proportion of missing genotypes, and a certain amount of naturally missing genotypes has to be allowed along with artificially introduced ones (see for instance \cite{rutkoski_imputation_2013}). We extracted four datasets from the overall data that had a maximum of 10\%, 20\%, 40\% or 70\% missing data per marker. For each of these datasets, we randomly introduced an additional 1\%, 5\%, 10\% or 20\% missing genotypes, on which imputation accuracy could be measured. Table~\ref{tab:missing_rates} reports the number of markers and average missing rate (before introducing artificial missing genotypes) for the four call-rate thresholds. This procedure produced 16 combinations for each population: 4 call rate datasets $\times$ 4 levels of artificial missing markers. On each combination, we applied the 5 described imputation methods. Rice data were analysed as a whole and by each of the 12 chromosomes separately. \\
For rice data only, the position of the SNPs on the genome was known. It was therefore possible to test the imputation performance with and without this information. Among the proposed imputation methods, only Beagle can use SNP position in the imputation process. We therefore applied Beagle both to SNP ordered per chromosome and along the entire genome and to SNP that were randomly reshuffled (so as to break the linkage between them). \\
All of the above led to a total of 1408 (1200 + 208 for reshuffled rice data with Beagle) imputation analyses.

@@

\subsection{Imputation accuracy and efficiency assessment}
\label{sec:imputation_accuracy_and_efficiency_assessment}
We used the artificial missing genotypes to measure the performance of the five imputation methods. The rice dataset contained only two genotype classes: AA (homozygous for the major allele) and BB (homozygous for the minor allele). Both alfalfa datasets contained also the third genotype classes AB (heterozygous, pooling the three possible heterozygotes AAAB, AABB, ABBB).\\
For each experiment we measured the overall imputation accuracy and the imputation accuracy within each genotype class. The imputation accuracy was computed as the number of missing data correctly imputed divided by the total number of artificially missing data (proportion of correct imputations):

@@

where $I()$ is an indicator function that returns 1 if the imputed ($\hat{g}$) and true ($g$) genotypes are equal, 0 otherwise. We obtained four accuracy measures for alfalfa (three genotype classes plus total accuracy) and three for rice (two genotype classes plus total accuracy).\\
For each imputation method, also the total computation time needed to complete the analyses was measured, in order to get an idea of their relative performance. The measured time equals the total dedicated CPU time in case of single thread execution, and corresponds to a fraction of it in case of algorithm supporting multi thread execution. In our experiments only RFI implemented parallel execution, and it was configured to use 10 CPUs, thus all the reported times are to be considered as time elapsed while employing ten times more computational resources than the other algorithms.\\
To ensure consistency in the experimental conditions, all analysis were run on the same computing platform: a multi-node cluster with 64 CPUs AMD Opteron(tm) 2400 MHz and 256GB RAM for each node.

\subsection{Software}
\label{sec:software}
Data handling, editing and preparation, summary of results and plots were all performed using the open-source environment for statistical programming R \cite{r_core_team_r:_2014}. All imputation methods, except Beagle, were implemented in R: MNI directly in base R; KNNI using the function \emph{knncatimpute()} from the Scrime package \cite{schwender_scrime:_2013}; SVDI with the R package ``bcv'' \cite{perry_bcv:_2009}; RFI using the ``MissForest'' \cite{stekhoven_missforestnon-parametric_2012} R package (with parameters: ntree=100, maxiter=10, parallelize='variables'). The ``Beagle'' localized haplotype clustering imputation method was implemented with the Beagle software version 3.3.2 \cite{browning_rapid_2007}.

>>>>

\subsection{Imputation methods}
\label{sec:imputation_methods}
We considered six imputation methods: Mean Value Imputation (MNI), K-Nearest Neighbour Imputation (KNNI), Random Forest Imputation (RFI), Singular Value Decomposition Imputation (SVDI), Localized Haplotype Clustering Imputation (``Beagle'') and the ``FILLIN'' algorithm from the ``Tassel'' programming suite. For all algorithms, we imputed a $m\times n$ matrix of m individuals and n markers whose data points, defined in the set \{0,1,2,NA\}, represented the three possible genotypes (AA, AB, and BB) and the missing value, respectively. 
\paragraph{MNI}
\label{par:MNI}
: in Mean Value Imputation each missing data point was simply replaced by the mean of the non-missing values for that marker, then discretized to the closest value in \{0,1,2\}.
\paragraph{KNNI}
\label{par:KNNI}
: in K-Nearest Neighbors Imputation, missing data points were imputed based on the weighted average of the K closest markers \cite{troyanskaya_missing_2001}. This K-neighbourhood was defined by the Simple Matching Coefficient distance function \cite{schwender_statistical_2007}, specifically designed for categorical data. $K=4$ was used in KNNI, and neighbour values were weighted by the reciprocal of their distance from the data point to be imputed.
\paragraph{SVDI}
\label{par:SVDI}
: Singular Value Decomposition (SVD) imputation was based on the following factorization of the genotype matrix M (m markers, n individuals):

@@

\paragraph{FILLIN}
The ``Fast Inbred Line Library ImputatioN'' (``FILLIN'') is an imputation method optimised for inbred populations implemented in the ``Tassel'' programming suite \cite{swarts_novel_2014}. FILLIN is based on haplotype reconstruction around recombination break points; haplotypes are then clustered per genotype similarity using the Hamming distance function. This information is eventually used to impute the target locus in an iterative approach that attemps, through a Markov Chain MonteCarlo (MCMC) process, to maximise the likelihood of the observed SNP calls given the unobserved imputed genotypes.

\subsection{Imputation procedure}
\label{sec:imputation_procedure}
In order to assess the imputation performance of the different methods, we introduced increasing proportions of artificial missing genotypes in the data sets. These were then imputed with the various algorithms, measuring the resulting imputation accuracy and computation time. We extracted four datasets from the overall data that had a maximum of 10\%, 20\%, 40\% or 70\% missing data per marker. For each of these datasets, we randomly introduced an additional 1\%, 5\%, 10\% or 20\% missing genotypes, on which imputation accuracy could be measured. Table~\ref{tab:missing_rates} reports the number of markers and average missing rate (before introducing artificial missing genotypes) for the four call-rate thresholds. This procedure produced 16 combinations for each population: 4 call rate datasets $\times$ 4 levels of artificial missing markers. On each combination, we applied the 6 described imputation methods. Rice data were analysed as a whole and by each of the 12 chromosomes separately. In absence of alfalfa genome, the Medicago Truncatula genome was used as reference for those algorithms (Beagle and FILLIN) requiring markers positions.

To further investigate the effect of presence (or absence) of genome information we generated ``reshuffled'' datasets where marker positions were randomly assigned inside each chromosome. This way, the linkage disequilibrium between markers is broken and the relative performance decrease can be directly assessed. The reshuffled datasets were tested only on Beagle and FILLIN.

@@

\subsection{Imputation accuracy and efficiency assessment}
\label{sec:imputation_accuracy_and_efficiency_assessment}
We used the artificial missing genotypes to measure the performance of the six imputation methods. The rice dataset contained only two genotype classes: AA (homozygous for the major allele) and BB (homozygous for the minor allele). Both alfalfa datasets contained also the third genotype classes AB (heterozygous, pooling the three possible heterozygotes AAAB, AABB, ABBB).\\
For each experiment we measured the overall imputation accuracy and the imputation accuracy within each genotype class. The imputation accuracy was computed as the number of missing data correctly imputed divided by the total number of artificially missing data (proportion of correct imputations):

@@

where $I()$ is an indicator function that returns 1 if the imputed ($\hat{g}$) and true ($g$) genotypes are equal, 0 otherwise. We obtained four accuracy measures for alfalfa and three for rice.\\
For each imputation method, the total computation time needed to complete the analyses was measured as an indicator of their relative performance. The measured time equals the total dedicated CPU time in case of single thread execution, and corresponds to a fraction of it in case of algorithm supporting multi thread execution. In our experiments only RFI implemented parallel execution, and it was configured to use 10 CPUs, thus all the reported times are to be considered as time elapsed while employing ten times more computational resources than the other algorithms.\\
To ensure consistency in the experimental conditions, all analysis were run on the same computing platform: a multi-node cluster with 64 CPUs AMD Opteron(tm) 2400 MHz and 256GB RAM for each node.

\subsection{Software}
\label{sec:software}
Data handling, editing and preparation, summary of results and plots were all performed using the open-source environment for statistical programming R \cite{r_core_team_r:_2014}. All imputation methods except Beagle and FILLIN were implemented in R: MNI directly in base R; KNNI using the function \emph{knncatimpute()} from the Scrime package \cite{schwender_scrime:_2013}; SVDI with the R package ``bcv'' \cite{perry_bcv:_2009}; RFI using the ``MissForest'' \cite{stekhoven_missforestnon-parametric_2012} R package (with parameters: ntree=100, maxiter=10, parallelize='variables'). The ``Beagle'' localized haplotype clustering imputation method was implemented with the Beagle software version 3.3.2 \cite{browning_rapid_2007}. The FILLIN algorithm is implemented using the Tassel CLI plugin \emph{-FILLINFindHaplotypesPlugin} followed by \emph{-FILLINImputationPlugin}. The latter allowed accuracy measurements through \emph{-accuracy} and \emph{-propSitesMask} options.
